Subject-level developmental relationships

### 2024-08-30 2_developmental_analysis.qmd
analyses

>>>>

# Variability

```{r}
d_trial |>
  group_by(administration_id, age) |>
  summarise (sd_rt = sd(rt, na.rm=TRUE), 
             sd_acc = sd(long_window_accuracy)) |>
  pivot_wider(
ggplot(aes(x = age, y = sd_rt)) + 
  geom_point(alpha = .05) + 
  geom_smooth() +
  geom_smooth(method = "lm", col = "green") +
  geom_hline(yintercept = .5, lty = 2) + 
  scale_x_log10() 
```

### 2024-08-30 3_developmental_relationships.qmd
analyses

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(source(here("helper","common.R")))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
```

This is broken due to some repeated fields.

@@

```{r}
m_rt <- mean(d_sub_cdi$rt, na.rm=TRUE)
sd_rt <- sd(d_sub_cdi$rt, na.rm=TRUE)
d_sub_cdi$factor_rt <- cut(d_sub_cdi$rt, 
                           breaks = quantile(d_sub_cdi$rt, 
                                             probs = c(0, .25, .5, .75, 1), 
                                             na.rm= TRUE))

ggplot(filter(d_sub_cdi, !is.na(rt), !is.na(factor_rt)),
       aes(x = log_age, 
                      y = long_window_accuracy, 
                      col = factor_rt)) +
  geom_point(alpha = .3) +
  geom_smooth() +
  geom_hline(yintercept = .5, lty = 2)+
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "RT quantile")

@@

ggplot(d_sub_cdi, 
       aes(x = rt, 
                      y = long_window_accuracy, 
                      col = age_group)) +
  geom_point(alpha = .3) +
  geom_smooth() +
  geom_hline(yintercept = .5, lty = 2)+
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "age group")
```

<!-- By dataset.  -->

<!-- ```{r} -->
<!-- ggplot(acc_rt, aes(x = log_rt, y = elogit, col = log_age)) + -->
<!--   geom_point(alpha = .3) + -->
<!--   geom_smooth(method = "lm") + -->
<!--   geom_hline(yintercept = .5, lty = 2) + -->
<!--   facet_wrap(~dataset_name) +  -->
<!--   xlab("Mean log RT") +  -->
<!--   ylab("Mean elogit accuracy") -->

<!-- ``` -->

## Dimensionality reduction

Make into matrices.

```{r control-for-age}
d_sub_cdi_mat <- d_sub_cdi |>
  ungroup() |>
  select(age, rt,  long_window_accuracy, prod) |>
  mutate(rt = scale(rt / age)[,1],
         long_window_accuracy = scale(long_window_accuracy / age)[,1], 
         prod = scale(prod / age)[,1]) |>
  select(-age)

@@

acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)

@@

Controlling for age, the first component (~49% of variance) represents high production, fast RT, but low LWL accuracy.
The second component (~31% of variance) captures slow RT, low accuracy, and moderate production, and the third component (~20%) captures fast RT, low accuracy, and low production).

@@

```{r}
d_sub_cdi_mat <- d_sub_cdi |>
  ungroup() |>
  select(log_age, rt,  long_window_accuracy, prod) |>
  mutate(log_age = scale(log_age)[,1], 
         rt = scale(rt)[,1],
         long_window_accuracy = scale(long_window_accuracy)[,1], 
         prod = scale(prod)[,1])

d_sub_cdi_mat <- d_sub_cdi_mat |>
  filter(complete.cases(d_sub_cdi_mat)) |>
  as.matrix() 
```

@@

```{r}
# scree(d_sub_cdi_mat, pc=FALSE)  # Use pc=FALSE for factor analysis

# install.packages("psych")
mod1 <- fa(d_sub_cdi_mat, 1)
mod2 <- fa(d_sub_cdi_mat, 2)

@@

Let's do principal components analysis over the scaled variables.

acc_rt_prc <- prcomp(d_sub_cdi_mat, 2)

There is a first component capturing ~50% of the variance that is low accuracy, low production, slower, and younger.
A second component capturing ~30% of the variance is higher accuracy, low production, fast RT, and slightly young.
A third component (~16%) is older, highly accurate, low production, and somewhat slow RT.

>>>>

```{r}
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(psych))
suppressPackageStartupMessages(source(here("helper","common.R")))
d_sub <- readRDS(here("cached_intermediates","1_d_sub.Rds"))
load(here("..", "peekbank-method", "cached_intermediates", "1_cdi_subjects.Rds"))
```

This is very slightly broken due to some repeated fields. We use `values_fn = mean` to suppress them.

@@

```{r}
m_rt <- mean(d_sub_cdi$rt, na.rm=TRUE)
sd_rt <- sd(d_sub_cdi$rt, na.rm=TRUE)
d_sub_cdi$factor_rt <- cut(d_sub_cdi$rt, 
                           breaks = quantile(d_sub_cdi$rt, 
                                             probs = c(0, .25, .5, .75, 1), 
                                             na.rm= TRUE), 
                           labels = c("0-25", "25-50", "50-75", "75-1"))

ggplot(filter(d_sub_cdi, !is.na(rt), !is.na(factor_rt)),
       aes(x = age, 
           y = long_window_accuracy, 
           col = factor_rt)) +
  geom_point(alpha = .3) +
  geom_smooth() +
  scale_x_log10() + 
  geom_hline(yintercept = .5, lty = 2)+
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "RT quantile") + 
  xlab("Age (log months)") + 
  ylab("Accuracy (long window)") + 
  theme(legend.position = "bottom")

@@

ggplot(filter(d_sub_cdi, !is.na(age_group)), 
       aes(x = rt, 
           y = long_window_accuracy, 
           col = age_group)) +
  geom_point(alpha = .3) +
  geom_smooth() +
  geom_hline(yintercept = .5, lty = 2)+
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "age group")
```

By dataset.

```{r}
datasets_with_age_variation <- d_trial |>
  group_by(dataset_name) |>
  summarise(age_range = max(age) - min(age)) |>
  filter(age_range >= 6)

ggplot(filter(d_sub_cdi, 
              !is.na(rt), !is.na(factor_rt), 
              dataset_name %in% datasets_with_age_variation$dataset_name),
       aes(x = age, 
           y = long_window_accuracy, 
           col = factor_rt)) +
  geom_point(alpha = .3) +
  geom_smooth(se= FALSE, method = "lm") +
  scale_x_log10() + 
  geom_hline(yintercept = .5, lty = 2)+
  viridis::scale_color_viridis(discrete = TRUE, 
                               name = "RT quantile") +
  facet_wrap(~dataset_name) + 
  xlab("Age (log months)") + 
  ylab("Accuracy (long window)")

# Dimensionality reduction

## Production

Make into matrices and look at PCA

```{r}
d_sub_cdi_mat <- d_sub_cdi |>
  ungroup() |>
  select(log_age, rt,  long_window_accuracy, prod)

@@

acc_rt_prc <- prcomp(x = d_sub_cdi_mat, scale = TRUE)

@@

@@

@@

```{r}
mod1 <- fa(d_sub_cdi_mat, 1)
mod2 <- fa(d_sub_cdi_mat, 2)
```

Scree plot

```{r}
fa.parallel(d_sub_cdi_mat)
```

## Comprehension

```{r}
d_sub_cdi_mat_comp <- d_sub_cdi |>
  ungroup() |>
  select(log_age, rt,  long_window_accuracy, comp)

d_sub_cdi_mat_comp <- d_sub_cdi_mat_comp |>
  filter(complete.cases(d_sub_cdi_mat_comp)) |>
  as.matrix()

acc_rt_prc_comp <- prcomp(x = d_sub_cdi_mat_comp, scale = TRUE)

acc_rt_prc_comp
summary(acc_rt_prc_comp)
ggbiplot::ggbiplot(acc_rt_prc_comp, alpha = .1)
```

```{r}
mod1_comp <- fa(d_sub_cdi_mat_comp, 1)
mod2_comp <- fa(d_sub_cdi_mat_comp, 2)
```

```{r}
fa.parallel(d_sub_cdi_mat_comp)
```

# Longitudinal data

```{r}
longitudinal <- d_sub_cdi |>
  group_by(dataset_name, subject_id) |>
  count() |>
  filter(n > 1)

d_sub_long <- filter(d_sub_cdi, subject_id %in% longitudinal$subject_id)

d_sub_wide <- d_sub_long |>
  group_by(subject_id, 
```

```{r}
library(lavaan)

@@
